---
title: "DadaTrial"
output: pdf_document
date: "2026-01-16"
---

# TODO:
- Figure out what is needed for package installation when running as script
- Adjust filtering parameters to meet our goals


```{r package load}
library(dada2); packageVersion("dada2")
```

```{r path set}
inpath <- snakemake@input[[]]
outpath <- "."
list.files(inpath)
```


```{r read categories}
# read file names for forward and reverse
fnFs <- sort(list.files(inpath, pattern=".R1.fastq", full.names = TRUE))

# extracted sample names from forward reads
sample.names <- sapply(strsplit(basename(fnFs), "\\."), `[`, 2)
```

```{r plotting readqual}
#forward reads (first 6)
plotQualityProfile(fnFs[1:6])
```

```{r filtered path setup}
filtFs <- file.path(outpath, "filtered", paste0(sample.names, "_R1_filt.fastq.gz"))

names(filtFs) <- sample.names
```

```{r filtering and trimming}
# Chunking forward reads in to processing batches to overcome memory use issues
chunk_size <- 5
chunk_idx <- split(seq_along(fnFs), ceiling(seq_along(fnFs) / chunk_size))
outs <- vector("list", length(chunk_idx))

for (i in seq_along(chunk_idx)) {
  j <- chunk_idx[[i]]
  message(sprintf('Chunk #%d (%d files)', i, length(j)))
  outs[[i]] <- filterAndTrim(fnFs[j], filtFs[j],
                     truncLen = 190,
                     trimLeft = 7,
                     maxN = 0, maxEE = 2, truncQ = 2, rm.phix = TRUE,
                     compress = FALSE, multithread = FALSE)
}
out <- do.call(rbind, outs)

head(out)
```
```{r sample count sanity check}
length(filtFs)
sum(file.exists(filtFs))
table(file.size(filtFs) == 0)
```

```{r Learn Error Rates}
errF <- learnErrors(filtFs, multithread = TRUE)
plotErrors(errF, nominalQ = TRUE)
```
- learnErrors will stop taking base reads from samples once it reaches a sufficiently large number to work from for error modeling 
- hence the much lower number of samples used for error modeling compared to what is available when dealing with samples of high read count
- error plots are not looking good, (error rate v consensus quality score seems to be relatively flat for all possible substitutions)
```{r sample inference}
dadaFs <- dada(filtFs, err=errF, multithread=TRUE)
```

```{r Sequence Table Construction}
seqtab <- makeSequenceTable(dadaFs)
dim(seqtab)
table(nchar(getSequences(seqtab)))
```

```{r Read Count Tracking}
getN <- function(x) sum(getUniques(x))
track <- cbind(out, sapply(dadaFs, getN), rowSums(seqtab))
colnames(track) <- c("input", "filtered", "denoised", "seqtable")
rownames(track) <- sample.names
head(track)
```
